Determinism, Complexity, and Predictability in Computer Performance

### 2014-04-01 related.tex
some edits, and making comments red

We build on the notion of \emph{redundancy} that was introduced on
page~\pageref{page:redundancy}, which formally quantifies how
information propagates forward through a time series:
% how much information prior observations of a system lend to future
% values
i.e., the mutual information between the past $n$ observations and the
current one.
% took this out because DCE hasn't been introduced yet: 
% 
% For example,
% this would be the amount of information that delay coordinate
% embedding captures.
The redundancy of i.i.d. random processes, for instance, is zero,
since all observations in such a process are independent of one
another.  On the other hand, deterministic systems---including chaotic
ones---have maximal redundancy and thus can be perfectly predicted if
observed for long enough~\cite{weigend-book}.  In practice, it is
quite difficult to estimate the redundancy of an arbitrary,
real-valued time series.  Doing so requires knowing either the
Kolmolgorov-Sinai entropy or the values of all positive Lyapunov
exponents of the system.  Both of these calculations are
difficult---the latter particularly so if the data are very noisy or
the generating system is stochastic.

>>>>

We build on the notion of \emph{redundancy} that was introduced on
page~\pageref{page:redundancy}, which formally quantifies how
information propagates forward through a time series:
% how much information prior observations of a system lend to future
% values
i.e., the mutual information between the past $n$ observations and the
current one.
% took this out because DCE hasn't been introduced yet: 
% 
% For example,
% this would be the amount of information that delay coordinate
% embedding captures.
The redundancy of i.i.d. random processes, for instance, is zero,
since all observations in such a process are independent of one
another.  On the other hand, deterministic systems---including chaotic
ones---have high redundancy which is maximal in the infinite limit, 
and thus can be perfectly predicted if observed for long 
enough~\cite{weigend-book}.  In practice, it is
quite difficult to estimate the redundancy of an arbitrary,
real-valued time series.  Doing so requires knowing either the
Kolmogorov-Sinai entropy or the values of all positive Lyapunov
exponents of the system.  Both of these calculations are
difficult---the latter particularly so if the data are very noisy or
the generating system is stochastic.

### 2014-04-01 results.tex
some edits, and making comments red

The experiments below involve three different prediction methods
applied to time-series data from eight different systems: {\tt
  col\_major}, {\tt 403.gcc}, and the six different regimes of the
{\tt dgesdd} signal in Figure~\ref{fig:svd-ts-colored}.  The objective of these
experiments was to explore how prediction accuracy is related to WPE,
and how that relationship depends on the generating process and the
prediction method.  Working from the first 90\% of each signal, we
generated a prediction of the last 10\% using the \naive, ARIMA, and
LMA prediction methods, as described in Section~\ref{sec:accuracy},
then calculated the MASE value of those predictions.  We also
calculated the WPE of each time series using a wordlength of six, as
described in Section~\ref{sec:meaComplex}.  In order to assess the
run-to-run variability of these results, we repeated all of these
calculations on 15 separate trials: i.e., 15 different runs of each
program.  Figure~\ref{fig:wpe_vs_mase_best} shows the {\sl best} of
these 45 predictions for each system: i.e., the lowest error over all
15 trials and all three methods for each of the eight programs.  The
WPE is plotted against the corresponding MASE value here in order to
bring out the correlation between these two quantities.
\begin{figure}[htbp]
  \centering
  \includegraphics[width=\columnwidth]{figs/prediction_vs_entropy}
  \caption{Weighted permutation entropy vs. mean absolute scaled error
    (MASE) of the best prediction of each system.
% 
% (across all trials and all prediction methods) 
% 
The dashed line is a least-squares linear fit of all the points except
{\tt dgesdd$_1$}, which we have excluded for reasons explained in the
text.  [[Ryan, is that correct?]]}
  \label{fig:wpe_vs_mase_best}
\end{figure}

@@

Figure~\ref{fig:wpe_vs_mase_all} shows WPE vs. MASE plots for all 360
experiments.
\begin{figure}
  \centering
%    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
%    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
%    \includegraphics[width=\columnwidth]{figs/LMA_prediction_vs_entropy}
  \includegraphics[width=\columnwidth]{figs/predictions_vs_entropy}
\caption{WPE vs. MASE for all trials, methods, and systems. \svdone, \svdthree, and \svdfive are omitted from the top plot for scale reasons; their MASE scores are $2.676 \pm 4.328$, $31.386 \pm 0.282$, and $20.870 \pm 0.192$ respectively. The dashed line is the same as in the previous figure. Numerical values, including means and standard deviations of the errors, can be found in Table~\ref{tab:error} in the appendix.
% 
%  [[All traces except those from \svd$_1$ lie above the line,
%      indicating that LMA is better suited prediction method for the
%      computer traces considered.]]  [[This comes {\bf long} before the
%      associated discussion in the text, and it's not what this figure
%      is trying to show here.]]
% 
}
    \label{fig:wpe_vs_mase_all}
\end{figure} 
There are 15 points in each cluster, one for each trial.  (The points
in Figure~\ref{fig:wpe_vs_mase_best} are the leftmost of the points
for the corresponding system in any of these three plots.)  The WPE
values do not vary very much across trials.  For most traces, the
variance in MASE scores is low as well, resulting in small, tight
clusters.  In some cases---ARIMA predictions of {\tt col\_major}, for
instance---the MASE variance is larger, which spreads out the clusters
horizontally.  The distribution of the cluster {\sl positions} is
quite different in the three plots: clusters of predictions generated
with the nonlinear LMA method are closer to the dashed line, whereas
the naive and ARIMA prediction clusters are more spread out [[need to
    revisit this wording when get the third image]].  This is a
geometric validation of the second conjecture in this paper, as
described in the following paragraphs.

Though \col is a very simple program, its dynamics are actually quite
complicated, as discussed in Section~\ref{sec:intro}.  Recall from the
graphical representations in the right-hand column of
Figure~\ref{fig:gcc_vs_col} that the \naive and ARIMA prediction
methods do not perform as well on this signal as the nonlinear LMA
method.  The MASE values of the \naive and ARIMA predictions are $0.57
\pm$ 0.001 and $0.59 \pm 0.21$, respectively, across all 15 trials.
That is, the errors produced by these methods are a little over half
as large as the errors produced by the random-walk method---a
primitive strategy that simply uses the current value as the
prediction.  However, the WPE value for the \col trials is $0.51 \pm
0.003$, which is in the center of the complexity spectrum described in
Section~\ref{sec:intro}.  [[This suggests that roughly half of the
    signal is predictive structure.]]  [[I'm still uncomfortable with
    that phrasing; let's revisit it after we have the terminology and
    definitions hammered out.]]

@@

Note that the \col points are clustered in LMA but spread out
horizontally in ARIMA, far to the right of the dashed line (i.e., much
worse than the other methods).  [[We need to come up with a hypothesis
    about this spread and skew, even if it's totally tentative.  We
    may want to merge this discussion into one of the other \col
    paragraphs---or perhaps put it with the \svdone discussion later
    on?]]

The WPE of \svdfive ($0.677 \pm 0.006$) is higher than that of {\tt
  col\_major}.  This indicates that the rate of forward information
transfer of the underlying process is lower, but that time-series data
from this system still contain a significant amount of structure that
can, in theory, be leveraged to predict the future course of the time
series.  As mentioned above, though, in the discussion of {\tt
  col\_major}, the effectiveness of any prediction strategy will
depend on how well it leverages the available information; methods
that use linear models, for instance, may fail when applied to
nonlinear processes.
% 
% (This is the basis for the second conjecture above.)  
% 
The match between model and task is an issue in the \svdfive
experiments as well.  The MASE scores of the \naive and ARIMA
predictions for this system were $20.870 \pm 0.192$ and $2.370 \pm
0.051$, respectively: that is, 20.87 and 2.37 times worse than a
simple random walk forecast of the same signals\footnote{The \naive
  MASE score is large because the amount of variance in this signal is
  high, which makes guessing the mean a particularly bad strategy.
  The same thing is true of the \svdthree signal.[[Joshua, it isn't the variance it is the strength in the bimodal distribution and only predicting in one mode of the distribution...]]}.  Again, the
positions of these points on a WPE vs. MASE plot---significantly below
and to the right of the dashed line---should suggest to a practitioner
that there is more structure in this signal than the corresponding
method is able to leverage, and that a different method might do
better.  Indeed, for \svdfive, the LMA method produces a MASE score of
$ 0.718\pm 0.048 $ and a cluster of results that is much closer to the
dashed line on the WPE-MASE plot.  Again, this validates our second
conjecture: the LMA method can capture and reproduce the way in which
the \svdfive system processes information, but the \naive and ARIMA
prediction methods cannot.

The WPE of \gcc is much higher: $0.940 \pm 0.001$.  This system
transmits very little information forward in time and provides almost
no structure for prediction methods to work with.  [[What we see:
    \naive does the best, then LMA, then ARIMA.  Hypothesis: that that
    5\% of the structure is nonlinear, so LMA can use it better than
    ARIMA can.  \naive does better than LMA because it's filtering out
    the noise.]]

>>>>

The experiments below involve three different prediction methods
applied to time-series data from eight different systems: {\tt
  col\_major}, {\tt 403.gcc}, and the six different regimes of the
{\tt dgesdd} signal in Figure~\ref{fig:svd-ts-colored}.  The objective of these
experiments was to explore how prediction accuracy is related to WPE,
and how that relationship depends on the generating process and the
prediction method.  Working from the first 90\% of each signal, we
generated a prediction of the last 10\% using the \naive, ARIMA, and
LMA prediction methods, as described in Section~\ref{sec:accuracy},
then calculated the MASE value of those predictions.  We also
calculated the WPE of each time series using a wordlength of six, as
described in Section~\ref{sec:meaComplex}.  In order to assess the
run-to-run variability of these results, we repeated all of these
calculations on 15 separate trials: i.e., 15 different runs of each
program.  Figure~\ref{fig:wpe_vs_mase_best} shows the {\sl best} of
these 45 predictions for each system: i.e., the lowest error over all
15 trials and all three methods for each of the eight programs.  The
WPE is plotted against the corresponding MASE value here in order to
bring out the correlation between these two quantities.
\begin{figure}[htbp]
  \centering
  \includegraphics[width=\columnwidth]{figs/prediction_vs_entropy}
  \caption{Weighted permutation entropy vs. mean absolute scaled error
    (MASE) of the best prediction of each system.
% 
% (across all trials and all prediction methods) 
% 
The dashed line is a least-squares linear fit of all the points except
{\tt dgesdd$_1$}, which we have excluded for reasons explained in the
text.}
  \label{fig:wpe_vs_mase_best}
\end{figure}

@@

Figure~\ref{fig:wpe_vs_mase_all} shows WPE vs. MASE plots for all 360
experiments.
\begin{figure}
  \centering
%    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
%    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
%    \includegraphics[width=\columnwidth]{figs/LMA_prediction_vs_entropy}
  \includegraphics[width=\columnwidth]{figs/predictions_vs_entropy}
\caption{WPE vs. MASE for all trials, methods, and systems. \svdone, \svdthree, and \svdfive are omitted from the top plot for scale reasons; their MASE scores are $2.676 \pm 4.328$, $31.386 \pm 0.282$, and $20.870 \pm 0.192$ respectively. The dashed line is the same as in the previous figure. Numerical values, including means and standard deviations of the errors, can be found in Table~\ref{tab:error} in the appendix.
% 
%  [[All traces except those from \svd$_1$ lie above the line,
%      indicating that LMA is better suited prediction method for the
%      computer traces considered.]]  [[This comes {\bf long} before the
%      associated discussion in the text, and it's not what this figure
%      is trying to show here.]]
% 
}
    \label{fig:wpe_vs_mase_all}
\end{figure} 
There are 15 points in each cluster, one for each trial.  (The points
in Figure~\ref{fig:wpe_vs_mase_best} are the leftmost of the points
for the corresponding system in any of these three plots.)  The WPE
values do not vary very much across trials.  For most traces, the
variance in MASE scores is low as well, resulting in small, tight
clusters.  In some cases---ARIMA predictions of {\tt col\_major}, for
instance---the MASE variance is larger, which spreads out the clusters
horizontally.  The distribution of the cluster {\sl positions} is
quite different in the three plots: clusters of predictions generated
with the nonlinear LMA method are closer to the dashed line, whereas
the naive and ARIMA prediction clusters are more spread out {\color{red}[[need to
    revisit this wording when get the third image]]}.  This is a
geometric validation of the second conjecture in this paper, as
described in the following paragraphs.

Though \col is a very simple program, its dynamics are actually quite
complicated, as discussed in Section~\ref{sec:intro}.  Recall from the
graphical representations in the right-hand column of
Figure~\ref{fig:gcc_vs_col} that the \naive and ARIMA prediction
methods do not perform as well on this signal as the nonlinear LMA
method.  The MASE values of the \naive and ARIMA predictions are $0.57
\pm$ 0.001 and $0.59 \pm 0.21$, respectively, across all 15 trials.
That is, the errors produced by these methods are a little over half
as large as the errors produced by the random-walk method---a
primitive strategy that simply uses the current value as the
prediction.  However, the WPE value for the \col trials is $0.51 \pm
0.003$, which is in the center of the complexity spectrum described in
Section~\ref{sec:intro}.  {\color{red}[[This suggests that roughly half of the
    signal is predictive structure.]]  [[I'm still uncomfortable with
    that phrasing; let's revisit it after we have the terminology and
    definitions hammered out.]]  }

@@

Note that the \col points are clustered in LMA but spread out
horizontally in ARIMA, far to the right of the dashed line (i.e., much
worse than the other methods).  {\color{red}[[We need to come up with a hypothesis
    about this spread and skew, even if it's totally tentative.  We
    may want to merge this discussion into one of the other \col
    paragraphs---or perhaps put it with the \svdone discussion later
    on?]]}

The WPE of \svdfive ($0.677 \pm 0.006$) is higher than that of {\tt
  col\_major}.  This indicates that the rate of forward information
transfer of the underlying process is lower, but that time-series data
from this system still contain a significant amount of structure that
can, in theory, be leveraged to predict the future course of the time
series.  As mentioned above, though, in the discussion of {\tt
  col\_major}, the effectiveness of any prediction strategy will
depend on how well it leverages the available information; methods
that use linear models, for instance, may fail when applied to
nonlinear processes.
% 
% (This is the basis for the second conjecture above.)  
% 
The match between model and task is an issue in the \svdfive
experiments as well.  The MASE scores of the \naive and ARIMA
predictions for this system were $20.870 \pm 0.192$ and $2.370 \pm
0.051$, respectively: that is, 20.87 and 2.37 times worse than a
simple random walk forecast of the same signals\footnote{The \naive
  MASE score is large because the amount of variance in this signal is
  high, which makes guessing the mean a particularly bad strategy.
  The same thing is true of the \svdthree signal.{\color{red}[[Joshua, it isn't the variance it is the strength in the bimodal distribution and only predicting in one mode of the distribution...]]}}.  Again, the
positions of these points on a WPE vs. MASE plot---significantly below
and to the right of the dashed line---should suggest to a practitioner
that there is more structure in this signal than the corresponding
method is able to leverage, and that a different method might do
better.  Indeed, for \svdfive, the LMA method produces a MASE score of
$ 0.718\pm 0.048 $ and a cluster of results that is much closer to the
dashed line on the WPE-MASE plot.  Again, this validates our second
conjecture: the LMA method can capture and reproduce the way in which
the \svdfive system processes information, but the \naive and ARIMA
prediction methods cannot.

The WPE of \gcc is much higher: $0.940 \pm 0.001$.  This system
transmits very little information forward in time and provides almost
no structure for prediction methods to work with.  {\color{red}[[What we see:
    \naive does the best, then LMA, then ARIMA.  Hypothesis: that that
    5\% of the structure is nonlinear, so LMA can use it better than
    ARIMA can.  \naive does better than LMA because it's filtering out
    the noise.]]}
